---
title: "GWAS CATALOG"
output: html_document
date: "2024-05-03"
---

```{r setup, include=FALSE}
library(readr)
library(dplyr)
library(stringr)

```


# Load the csv file
Here the dataset (csv file from GWAS catalog) is loaded. Change the path accordingly.
```{r}
full_catalog <- readr::read_tsv("/home/maria/git/SOROLLA/SumStats/PAST_CSV/summary_statistics_table_export.tsv", num_threads = 1)
```

## Clean the data and only keep the rows that are relevant
We are only interested in European populations and data from GWAS, so we are going to exclude all the data that does not match these characteristics.
```{r}
#Filter by technology (only interested GWGA)
filtered_catalog <- full_catalog %>%
               filter(grepl("Genome-wide genotyping array", genotypingTechnologies, fixed = TRUE) &
                        !grepl("Exome-wide sequencing", genotypingTechnologies, fixed = TRUE) &
                        #Include and exclude ancestries
                        grepl("European", discoverySampleAncestry, fixed = TRUE) & 
                        !grepl("Asian", discoverySampleAncestry, fixed = TRUE) &
                        !grepl("Latin", discoverySampleAncestry, fixed = TRUE) &
                        !grepl("Hispanic", discoverySampleAncestry, fixed = TRUE) &
                        !grepl("African", discoverySampleAncestry, fixed = TRUE) &
                        !grepl("NR", discoverySampleAncestry, fixed = TRUE))
head(filtered_catalog)

#Extract numeric part and nationality, and add as new columns
filtered_catalog <- filtered_catalog %>%
               mutate(SampleSize = as.numeric(gsub("\\D", "", discoverySampleAncestry)),
                      Ancestry = gsub("\\d", "", discoverySampleAncestry))

#Filter the SampleSize to at least 10K
filtered_catalog <- filtered_catalog %>%
                filter(SampleSize >= 5000)

# Save filtered_catalog as a CSV file
write.csv(filtered_catalog, "/home/maria/git/SOROLLA/SumStats/FM_filtered_catalog.csv", row.names = FALSE)

# Confirmation message
cat("filtered_catalog has been saved as '/home/maria/git/SOROLLA/SumStats/FM_filtered_catalog.csv'")

```
Now create another dataframe filtering with the desired keywords.
(old)
terms <- c(
  "ADHD", "attention deficit", "hyperactivity", "ADD", "Anxiety", "Panic", "Autism", "Asperger", "Bipolar", "Depression", 
  "Manic", "Brain", "Cancer", "Neoplasm", "Carcinoma", "insanity", "cerebral", "central nervous", "malignancy", "psychosis", 
  "Pervasive Developmental Disorder", "breast", "colorectal", "colon", "anal", "rectal", "depression", "mood", "endometrial", 
  "endometrium", "endometriosis", "uterine", "uterus", "cervix", "cervical", "kidney", "liver", "hepatic", "ovarian", "ovaries", 
  "ovary", "pancreas", "pancreatic", "prostate", "testicular", "testis", "endocrine", "Schizophrenia", "psychotic", "thyroid", 
  "leukemia", "lymphoma", "melanoma", "sarcoma", "lung", "stomach", "stomacal", "gastrointestinal", "intestines", "gastric", 
  "cancerous", "tumor", "malignant", "benign", "metastasis", "gastrointestinal tract", "gastroenterology", "digestive system", 
  "esophagus", "stomach", "small intestine", "large intestine", "colon", "rectum", "anus", "liver", "gallbladder", "pancreas", 
  "bile ducts", "duodenum", "jejunum", "ileum", "cecum", "appendix", "rectum", "peritoneum", "lymph", "lymphatic", "hepatic", 
  "hepatocellular", "gut", "intestinal", "microbiota", "microbiome"
)
```{r}
# Define the terms to search for

terms <- c(
  "ADHD", "attention deficit", "hyperactivity", "ADD", "Anxiety", "Panic", "Autism", "Asperger", "Bipolar", "Depression", 
  "Manic", "Brain", "Cancer", "Neoplasm", "Carcinoma", "insanity", "cerebral", "central nervous", "malignancy", "psychosis", 
  "Pervasive Developmental Disorder", "breast", "colorectal", "colon", "anal", "rectal", "depression", "mood", "endometrial", 
  "endometrium", "endometriosis", "uterine", "uterus", "cervix", "cervical", "kidney", "liver", "hepatic", "ovarian", "ovaries", 
  "ovary", "pancreas", "pancreatic", "prostate", "testicular", "testis", "endocrine", "Schizophrenia", "psychotic", "thyroid", 
  "leukemia", "lymphoma", "melanoma", "sarcoma", "lung", "stomach", "stomacal", "gastrointestinal", "intestines", "gastric", 
  "cancerous", "tumor", "malignant", "benign", "metastasis", "gastrointestinal tract", "gastroenterology", "digestive system", 
  "esophagus", "stomach", "small intestine", "large intestine", "colon", "rectum", "anus", "liver", "gallbladder", "pancreas", 
  "bile ducts", "duodenum", "jejunum", "ileum", "cecum", "appendix", "rectum", "peritoneum", "lymph", "lymphatic", "hepatic", 
  "hepatocellular"
)

```

An addition to make the list more exhaustive:
```{r}
CNS <- c(
  "Alzheimer's disease", "Alzheimer disease", "Alzheimers", "Alzheimer", "Parkinson's disease", "Parkinson disease", "Parkinsons", "Parkinson",
  "Multiple sclerosis", "MS", "Amyotrophic lateral sclerosis", "ALS", "Lou Gehrig's disease", "Motor neuron disease", "MND", 
  "Epilepsy", "Seizure disorder", "Migraine", "Headache", "Cluster headache", "Tension headache", 
  "Dementia", "Frontotemporal dementia", "Vascular dementia", "Huntington's disease", "Huntington disease", "Huntingtons", "Huntington", 
  "Stroke", "Ischemic stroke", "Hemorrhagic stroke", "Brain injury", "Traumatic brain injury", "Encephalitis", "Meningitis", "Brain infection",
  "Neurodegeneration", "Neurodegenerative disease", "Cerebral palsy", "CP","Ataxia", "Cerebellar ataxia", "Spinocerebellar ataxia", 
  "Friedreich's ataxia", "Friedreichs ataxia", "Tourette syndrome", "Tremor", "Essential tremor", 
  "Myasthenia gravis", "Nerve disorder", "Neuropathy", "Peripheral neuropathy", "Neuromuscular disorder", "Neuromuscular disease", 
  "Brain tumor", "Benign brain tumor", "Malignant brain tumor", "Glioma", "Meningioma", "Astrocytoma", "Oligodendroglioma", "Ependymoma", 
  "Medulloblastoma", "Neuroblastoma", "Pituitary tumor", "Pituitary adenoma", "Schwannoma", "Vestibular schwannoma", "Acoustic neuroma", 
  "Neurofibroma", "Brainstem glioma", "Pontine glioma", "Spinal cord tumor", "Intramedullary tumor", "Extramedullary tumor", "Dystonia", 
  "Restless legs syndrome", "RLS", "Chronic traumatic encephalopathy", "CTE", "Spinal muscular atrophy", "SMA", "Muscular dystrophy", 
  "Duchenne muscular dystrophy", "DMD", "Becker muscular dystrophy", "BMD", "Neurofibromatosis", "NF", "NF1", "NF2", 
  "Neurocutaneous disorder", "Tuberous sclerosis", "Recklinghausen disease", "Guillain-Barré syndrome", "Guillain-Barre syndrome", 
  "Guillain-Barre", "Chronic inflammatory demyelinating polyneuropathy", "CIDP", "Charcot-Marie-Tooth disease", "Charcot-Marie-Tooth syndrome",
  "CMT", "Neuritis", "Encephalopathy", "Encephalomyelitis", "Central sleep apnea", "Narcolepsy", "Trigeminal neuralgia", "Bell's palsy", 
  "Bells palsy", "Rett syndrome", "Tay-Sachs disease", "Tay Sachs disease", "Tay-Sachs", "Tay Sachs", "Canavan disease", 
  "Krabbe disease", "Metachromatic leukodystrophy", "Prion disease", "Creutzfeldt-Jakob disease", "Creutzfeldt-Jakob", 
  "Normal pressure hydrocephalus", "NPH", "Intracranial hypertension", "Pseudotumor cerebri", "Chiari malformation","Dandy-Walker syndrome", 
  "Dandy Walker syndrome", "Tethered cord syndrome", "Aneurysm", "Arteriovenous malformation", "AVM", 
  "Cavernous malformation", "Cavernoma", "Hydrocephalus", "Anencephaly", "Spina bifida", "Encephalocele", "Holoprosencephaly", 
  "Lissencephaly", "Pachygyria", "Polymicrogyria", "Schizencephaly", "Microcephaly", "Hydranencephaly", 
  "Agenesis of the corpus callosum", "Dysgenesis of the corpus callosum", "Corpus callosum agenesis", 
  "Cerebellar hypoplasia", "Pontocerebellar hypoplasia", "Dandy-Walker malformation", "Dandy Walker malformation", 
  "Arachnoid cyst"
)


PSY <- c(
  "Schizophrenia", 
  "Bipolar disorder", "Bipolar", "Manic depression", 
  "Major depressive disorder", "MDD", "Depression", "Clinical depression", "Unipolar depression", 
  "Anxiety disorder", "Generalized anxiety disorder", "GAD", "Anxiety", 
  "Obsessive-compulsive disorder", "OCD", 
  "Post-traumatic stress disorder", "PTSD", "Post traumatic stress disorder", 
  "Attention-deficit/hyperactivity disorder", "ADHD", 
  "Autism spectrum disorder", "ASD", "Autism", 
  "Personality disorder", 
  "Borderline personality disorder", "BPD", 
  "Antisocial personality disorder", "APD", 
  "Paranoid personality disorder", 
  "Histrionic personality disorder", 
  "Narcissistic personality disorder", 
  "Schizoaffective disorder", 
  "Delusional disorder", 
  "Panic disorder", 
  "Agoraphobia", 
  "Social anxiety disorder", "Social phobia", 
  "Selective mutism", 
  "Psychosis", 
  "Hallucination", 
  "Delusion", 
  "Dissociative identity disorder", "Multiple personality disorder", "Dissociative disorder", 
  "Conversion disorder", 
  "Somatic symptom disorder", "Somatization disorder", 
  "Psychosomatic disorder", 
  "Neurocognitive disorder", "Cognitive disorder"
)


CAN <- c(
  "Breast cancer", "Breast carcinoma", "Breast tumor", 
  "Lung cancer", "Lung carcinoma", 
  "Prostate cancer", "Prostate carcinoma", 
  "Colorectal cancer", "Colon cancer", "Rectal cancer", 
  "Liver cancer", "Hepatocellular carcinoma", 
  "Stomach cancer", "Gastric cancer", 
  "Pancreatic cancer", "Pancreatic carcinoma", 
  "Ovarian cancer", "Ovarian carcinoma", 
  "Endometrial cancer", "Uterine cancer", 
  "Bladder cancer", "Bladder carcinoma", 
  "Kidney cancer", "Renal cancer", "Renal cell carcinoma", 
  "Esophageal cancer", "Esophageal carcinoma", 
  "Head and neck cancer", 
  "Thyroid cancer", "Thyroid carcinoma", 
  "Melanoma", "Skin cancer", "Cutaneous melanoma", 
  "Non-Hodgkin lymphoma", "NHL", 
  "Hodgkin lymphoma", "Hodgkin's disease", 
  "Leukemia", 
  "Acute lymphoblastic leukemia", "ALL", 
  "Chronic lymphocytic leukemia", "CLL", 
  "Acute myeloid leukemia", "AML", 
  "Chronic myeloid leukemia", "CML", 
  "Multiple myeloma", 
  "Testicular cancer", "Testicular tumor", "Testicular carcinoma", 
  "Penile cancer", 
  "Sarcoma", "Soft tissue sarcoma", 
  "Bone cancer", "Osteosarcoma", 
  "Brain cancer", "Central nervous system cancer", "CNS cancer", 
  "Neuroblastoma", 
  "Wilms' tumor", "Wilms tumor", 
  "Retinoblastoma", 
  "Hepatoblastoma", 
  "Choriocarcinoma", 
  "Germ cell tumor", "Germinoma", "Embryonal carcinoma", "Yolk sac tumor", 
  "Carcinoid tumor", 
  "Thymoma", 
  "Mesothelioma", 
  "Kaposi's sarcoma"
)


ADD <- c(
  "Angiosarcoma", 
  "Chondrosarcoma", 
  "Ewing sarcoma", 
  "Fibrosarcoma", 
  "Gastrointestinal stromal tumor", "GIST", 
  "Histiocytoma", 
  "Liposarcoma", 
  "Melanotic neuroectodermal tumor", 
  "Myxofibrosarcoma", 
  "Neuroendocrine tumor", 
  "Osteosarcoma", 
  "Pheochromocytoma", 
  "Salivary gland cancer", 
  "Testicular germ cell tumor", 
  "Thyroid carcinoma", 
  "Uveal melanoma", 
  "Vulvar cancer", 
  "Appendix cancer", 
  "Biliary tract cancer", 
  "Cervical cancer", 
  "Cholangiocarcinoma", 
  "Endometrial carcinoma", 
  "Epithelial ovarian cancer", 
  "Gestational trophoblastic disease", 
  "Malignant pleural mesothelioma", 
  "Neurofibrosarcoma", 
  "Pleomorphic sarcoma", 
  "Primitive neuroectodermal tumor", "PNET", 
  "Reticulum cell sarcoma", 
  "Sinonasal cancer", 
  "Synovial sarcoma", 
  "Thyroid papillary carcinoma", 
  "Thyroid follicular carcinoma"
)


ALL <- c(CNS,PSY,CAN,ADD)
complete <- c(terms, ALL)
```



```{r}
# Function to check if any term is present in a string (case-insensitive)
check_terms <- function(string) {
  # Convert both the string and terms to lower case for case-insensitive comparison
  string <- as.character(string)
  terms_lower <- tolower(complete)
  # Check if any term is present in the string
  any(grepl(paste(terms_lower, collapse="|"), string, ignore.case=TRUE))
}

# Filter dataframe to rows where any of the terms are present in "reportedTrait" or "efoTraits" (corrected column name)
matches <- sapply(1:nrow(filtered_catalog), function(i) {
  check_terms(filtered_catalog[i, "reportedTrait"]) || check_terms(filtered_catalog[i, "efoTraits"])
})

words_filtered_catalog <- filtered_catalog[matches, ]
print(words_filtered_catalog)

```

```{r}
# Save filtered_catalog as a CSV file
write.csv(words_filtered_catalog, "/home/maria/git/SOROLLA/SumStats/FM_words_filtered_catalog.csv", row.names = FALSE)

# Confirmation message
cat("words_filtered_catalog has been saved as '/home/maria/git/SOROLLA/SumStats/FM_words_filtered_catalog.csv'")
```

```{r}
df <- read.csv("/home/maria/git/SOROLLA/SumStats/FM_words_filtered_catalog.csv")
```


```{r}
# Function to extract numbers and update Nca_val or Nco_val
extract_and_update <- function(description) {
  # Remove commas from the description
  description <- gsub(",", "", description)
  
  Nca_val <- ifelse(grepl("cases", description, ignore.case = TRUE),
                    as.numeric(str_extract(description, "\\d+(?=\\s*European ancestry cases)")),
                    NA)
  Nco_val <- ifelse(grepl("controls", description, ignore.case = TRUE),
                    as.numeric(str_extract(description, "\\d+(?=\\s*European ancestry controls)")),
                    NA)
  
  return(data.frame(Nca_val = Nca_val, Nco_val = Nco_val))
}

# Apply function to populate Nca_val and Nco_val columns
df <- df %>%
  mutate(extracted_vals = extract_and_update(initialSampleDescription),
         Nca_val = extracted_vals$Nca_val,
         Nco_val = extracted_vals$Nco_val) %>%
  select(-extracted_vals)

```

```{r}
write.csv(df, "/home/maria/git/SOROLLA/SumStats/FM_words_filtered_catalog.csv", row.names = FALSE)

```


# Now we filter through the ones that I had already annotated somehow (because I just want to add new ones) and also through the Fabian ones.

```{r}
#Load annotated dataset with Yes/not enough N...
annotated <- read.csv("/home/maria/git/SOROLLA/SumStats/PAST_CSV/annotated_past_catalog.csv")
Fabian <- read_csv("/home/maria/git/HDL-scripting/R/Datasets-info.csv")
```

remove_list <- annotated %>%
  filter(SelectionFilter %in% c("N not enough", "Unsure", "not specific", "Not relevant", "Not SNP", "MICROBIOME")) %>%
  pull(accessionId)

words_filtered_from_annotated <- df %>%
  # Keep rows where AccessionId is not in the list
  filter(!(accessionId %in% remove_list))

words_filtered_from_annotated <- words_filtered_from_annotated %>%
  # Keep rows where AccessionId is not in the list
  filter(!(efoTraits %in% complete))

```{r}
list_ids_remove <- annotated$accessionId
df_clean <- df %>%
  filter(!(accessionId %in% list_ids_remove))
```



```{r}
df_clean <- df_clean %>%
  # Keep rows where AccessionId is not in the list
  filter(!(efoTraits == "blood protein measurement"))

df_clean <- df_clean %>%
  # Keep rows where AccessionId is not in the list
  filter(!(efoTraits == "intestinal infectious diseases"))

df_clean <- df_clean[!grepl("measurement", df_clean$efoTraits), ]

```

```{r}
write.csv(df_clean, "/home/maria/git/SOROLLA/SumStats/PAST_CSV/FM_words_filtered_catalog_clean.csv", row.names = FALSE)

```

