---
title: "supergnova_visualize"
output: html_document
date: "2024-09-25"
---


# Supergnova results visualization
The supergnova software is best at estimating covariance over correlation. However, it is a good estimate of what regions could be related to the results from ldsc.
-------------------------------------------------------------

The colums we are going to see are:

- chr: The chromosome.
- start: The start position of the genomic region from the input genome partition file.
- end: The end position of the genomic region from the input genome partiition file.
- rho: The estimation of local genetic covariance.
- corr: The estimation of local genetic correlation.
- h2_1: The estimation of local heritability of the first trait by method of moments.
- h2_2: The estimation of local heritability of the second trait by method of moments.
- var: The variance of the estimation of local genetic covariance.
- p: The p value of local genetic covariance.
- m: The number of SNPs involved in the estimation of local genetic covariance in the genomic region.

We are mainly interested in the start and end position in the chromosome to figure out what gene is responsible for the results. And then the correlation variable as well as heritability.

-------------------------------------------------------------
So the tasks for this notebook that will be later turned into an automatised script to obtain clean csvs and figures if necessary, as well as the imputation of the genes are:

1. Clean missing values from corr column.
2. Impute genes involved in results.
3. Visualise the unique genes and infer pathways related to them.

-------------------------------------------------------------

```{r}
#Libraries
library(ggplot2)
library(dplyr)  
library(tidyr)  
library(ggplot2)
```
## Obtain data:
Here we would need to automatise it to extract the disease names, and write in a text file:
*Before cleaning*:
1. Number of rows
2. Number of missing data from corr
*After cleaning*:
1. Number of remaining rows.

##### Read file
```{r}
supergnova <- read.table("/home/maria/git/SOROLLA/Results/Supergnovatest.txt", sep=" ", header=TRUE)
print(supergnova)
print(dim(supergnova))
```


#### Remove rows with missing values in the "corr" column
```{r}
supergnova_clean <- supergnova[complete.cases(supergnova$corr), ]
```

#### Sort the cleaned data frame by "corr" in ascending order
```{r}
ordered_clean_supergnova <- supergnova_clean[order(supergnova_clean$corr), ]
```


```{r}
print(ordered_clean_supergnova)
print(dim(ordered_clean_supergnova))
```
```{r}
colnames(ordered_clean_supergnova)
```


## Error with values
We observe values much higher than -1 and 1. Which generally should not occur. For now we are going to remove them to visualise the ones we want to keep.

### Filter data based on correlation
```{r}
filtered_clean_supergnova <- ordered_clean_supergnova[ordered_clean_supergnova$corr > -1.01 & ordered_clean_supergnova$corr < 1.01, ]
```

## Plot distribution of SNPS in the chromosome
```{r}
chr_distribution <- ggplot(filtered_clean_supergnova, aes(x = chr)) +
  geom_bar(aes(fill = chr), stat = "count") +
  labs(title = "Chromosome Distribution", x = "Chromosome", y = "Frequency") +
  scale_x_discrete(breaks = seq(1, 22)) +
  theme_classic() +
  stat_count(geom = "text", aes(label = after_stat(count)), vjust = -0.5, ha = "center")
plot(chr_distribution)
```

# SNP count by chromosome plot
```{r} 
snp_count_plot <- ggplot(filtered_clean_supergnova, aes(x = chr, y = m, group = chr)) +
  geom_boxplot() +
  labs(title = "SNP Count by Chromosome", x = "Chromosome", y = "SNP Count") +
  scale_x_discrete(breaks = seq(1, 22)) +
  theme_classic()
plot(snp_count_plot)
```

# Retrieve the gene symbols
The concept is to use Biomart from Ensembl to retrieve the gene sign and then obtain the pathways that are related to said gene. First we want to know what genes represent the most correlation.

We connect to biomart
```{r}
mart <- useMart("ensembl")
mart <- useDataset("hsapiens_gene_ensembl", mart)
```

Iterate through the rows to add the genes to a list that will be then added as a column to our dataframe. There is an issue with timeout, so a batch size is determined (50-100 choose).

### Let's create a function.

```{r}
fetch_genes <- function(start_row, end_row, df, mart){
  #Initialize the empty list to collect the gene symbols
  gene_names <- list()
  #Create a loop that will iterate throught each row
  for (i in start_row:end_row){
    genes <- getBM(
      attributes = c("ensembl_gene_id", "external_gene_name", "start_position", "end_position", "chromosome_name"),
      filters = c("chromosome_name", "start", "end"),
      values = list(df$chr[i], df$start[i], df$end[i]),
      mart = mart
    )
    #Now append the retrieved genes to the list
    gene_names[[i]] <- genes$external_gene_name
  }
  return(gene_names)
}
```

The function is now created. Now we need to split into batches to run the function and then collect all the results into the df.

```{r}
batch_size <- 100

#Empty list of all the genes from the batches
all_gene_names <- list()
```

Now we loop using the batch size with the function we created to avoid timeout.
#### Loop iteration with function fetch_genes()
```{r}
for (start_row in seq(1, nrow(filtered_clean_supergnova), by = batch_size)) {
  end_row <- min(start_row + batch_size - 1, nrow(filtered_clean_supergnova))
  
  batch_gene_names <- fetch_genes(start_row, end_row, filtered_clean_supergnova, mart)
  
  # Store the result in the main list
  all_gene_names[start_row:end_row] <- batch_gene_names
}
```

Combine into a single column as a list (then we need to expand the df to have a column for each )
```{r}
filtered_clean_supergnova$gene_names <- sapply(all_gene_names, function(x) paste(x, collapse = ", "))
```


This is the initial code used which worked, but just in case to avoid the timeout we created the other function and loops.
```{r}
# gene_names <- list()
# 
# # Loop through each row in your DataFrame
# for (i in 1:nrow(filtered_clean_supergnova)) {
#   genes <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "start_position", "end_position", "chromosome_name"),
#                  filters = c("chromosome_name", "start", "end"),
#                  values = list(filtered_clean_supergnova$chr[i], filtered_clean_supergnova$start[i], filtered_clean_supergnova$end[i]),
#                  mart = mart)
#   
#   # Append the gene names to the list
#   gene_names[[i]] <- genes$external_gene_name
# }
# # 
# # Combine the gene names into a single column
# filtered_clean_supergnova$gene_names <- sapply(gene_names, function(x) paste(x, collapse = ", "))
```


```{r}
test1_genes <- getBM(attributes = c("ensembl_gene_id", "external_gene_name"),
                     filters = c("chromosome_name", "start", "end"),
                     values = list("12", 115329869, 117087376),
                     mart = dataset)
```

